I Compensate, therefore I Am \\ (accurate for DNN verification)

### 2024-01-27 Paper/proofs.tex
proof

Assume that there is a no (weakly) compensating pair of paths.

@@

Using the concept of sign, now we introduce the key lemma for Theorem \ref{th1}.

@@

		\item $\vx^{\sharp}$ is the input vector defined in the following way:
		\begin{itemize}
			\item $x^{\sharp}_{a_i}=\min(a_i)$ if $S(a_i)\in \{0,1\}$, and
			\item $x^{\sharp}_{a_i}=\max(a_i)$ otherwise, that is when $S(a_i)=-1$.
		\end{itemize}
	\end{itemize}

\end{definition}

@@

				Suppose $\mathcal{D}$ is a DNN with $\ReLU$ as the unique activation functions and with a unique output node $z$. 
				Let $X$ be the set of nodes $n$ such that there is a weak compensating path 
				passing by $n$ (except its source and target nodes).
				Then $\MILP_X$ computes the exact bounds $[\alpha,\beta]$ for $z$,
				i.e. $\alpha=\min(z)$ and $\beta=\max(z)$.
			\end{theorem}

@@

>>>>

Assume that there is no (weakly) compensating pair of paths.

@@

Using the concept of sign, we introduce input vectors $\vx^{*}, \vx^{\sharp}$:

@@

		\item $\vx^{\sharp}$ is the input vector defined in the following way:
		\begin{itemize}
			\item $x^{\sharp}_{a_i}=\min(a_i)$ if $S(a_i)\in \{0,1\}$, and
			\item $x^{\sharp}_{a_i}=\max(a_i)$ otherwise, that is when $S(a_i)=-1$.
		\end{itemize}
	\end{itemize}
\end{definition}

We can then prove the following key lemma for Theorem \ref{th1}.

@@

				Suppose $\mathcal{D}$ is a DNN with a unique output node $z$.
				Let $X$ be the set of nodes $n$ such that there is a weak compensating path 
				passing by $n$ (except its source and target nodes).
				Then $\MILP_X$ computes the exact bounds $[\alpha,\beta]$ for $z$,
				i.e. $\alpha=\min(z)$ and $\beta=\max(z)$.
			\end{theorem}

@@

\subsection{Revised Proof of Theorem \ref{th2}}

We assume without loss of generality that every node can be reached from at least one input with a path of non-zero weight (otherwise we just remove this node altogether).
We also assume without loss of generality that there is no node which can only reach $z$ with path of weight $0$, as such nodes have no contribution to the value $\val(z)$. We denote $m<n$ to mean that there exists a path with non zero weight from $m$ to $n$.

Further, we assume the fairly light {\em well-connected hypothesis} (H1) that the DNN is such that for every 2 (not necessarily disjoint) paths $\rho_1,\rho_2$ with non-zero weight with the same source $m$ and the same target $n$ at least 2 layers apart, there exists another path $\rho_3$ from $m$ to $n$ (with non zero-weight) totally disjoint from $\rho_1$ and $\rho_2$ (except at the source and target). This hypothesis is used to remove corner cases 
which will not happen in actual DNNs, which all are well-connected.

			We now show Theorem \ref{th2} under (H1).
			Let $X$ be the set of nodes on compensating paths with target $z$, but not as source or target.

			Let $Z$ be the subset of nodes $n$ of the DNN such that $S$ is defined on $n$ 
			wrt node $z$. As proved earlier, $Z$ is suffix closed, i.e. if $m<n<z$ with $m \in Z$, then $n \in Z$. We can partition $Z = Z_{+} \sqcup Z_{-}$, for positive and negative sign. $X$ can intersect both $Z_+$ and $Z_-$.
			We call $X_+ = Z_+ \cap X$ and $X_-=Z_- \cap X$, as well as 
			defined $Y_+,Y_-$ such that $Z_+ = X_+ \sqcup Y_+$ and $Z_- = X_- \sqcup Y_-$.
			We denote $Y=Y_+ \sqcup Y_-$.

\begin{lemma}
 Let $n \notin Z$. Then $n \in X$.
\end{lemma}

\begin{proof}
			node $n \notin Z$ implies the existence of two path $\rho_1,\rho_2$ (not necessarily disjoint) to $z$, with opposite sign. 
			Take an input node $a$ reaching $n$ with path $\rho$ with non zero weight.
			Then using (H1), we get a third path $\rho_3$ disjoint from $\rho \cdot \rho_1$ and from $\rho \cdot \rho_2$ with non-zero weight. Depending on the sign of $weight(\rho_3)$, it will form a compensating pair of path to $z$ either with $\rho \cdot \rho_1$ or with $\rho \cdot \rho_2$, hence $n$ will be in $X$. We call $X_*$ this subset $X$. \qed
\end{proof}

Hence, it means that $X \sqcup Y$ partitions the set of neurons in hidden layers.

  Hence $\rho_1$ must meet $\rho_2$ (that is before or at $y$).
  Their weights must be of opposite sign, otherwise we can inductively reapplied the same argument to their suffixes.
  Let $\rho_1=\rho_1'\rho_1''$.
  Both $\rho_1 \cdot \rho$ and $\rho_2' \cdot \rho_1'' \cdot \rho$
  are from $m$ to $z$ going through $y$, but with weights of opposite sign (as $\rho'_1$ and $\rho'_2$ have opposite sign).
  Now, by $(H1)$, there exists a path $\rho_3$ disjoint from $\rho_1 \cdot \rho$ and$\rho_2' \cdot \rho_1'' \cdot \rho$ from $m$ to $z$. It means that we have a pair of compensating paths from $m$ to $z$, and thus $y \in X$, a contradiction with $y \in Y_+$. \qed
\end{proof}

Hence, we obtain as corollary of both lemmas:

\begin{corollary}
\label{cor:simple}
Assume that $m < n$ with $n \in Y$.
Then $m \in Z$.
\end{corollary}

To prove Theorem \ref{th2}, we can then optimize $Y$ first on its own (considering the sign of every node): use the same argument in Section A, the proof of Theorem \ref{th1}, we can maximize all $n$ such that $S(n)=1$ and minimize all $n$ such that $S(n)=-1$ at the same time, i.e., with the same input values. And for every node $n$ in $Y$, the value computed is the same as actual lower and upper bounds of $n$.

Now we consider the contribution  (direct and indirect) of nodes in $Y$ to $z$ as biases. Let $z=\boldsymbol{bias}_z+\sum_{m\in Y} Appx(W_{mz}m)+\sum_{m\notin Y} \ReLU(W_{mz}m)$. Because we have pre-optimized nodes in $Y$, $\sum_{m\in Y} Appx(W_{mz}m)$ will be a constant in the following optimization, and its term $W_{mz}z$ can be regarded as a part of bias.

This is similar for nodes in other layers. Let $n\in X$ and $n=\boldsymbol{bias}_n+\sum_{m\in Y} Appx(W_{mn}m)+\sum_{m\notin Y} \ReLU(W_{mn}m)$; similarly, $\sum_{m\in Y}Appx(W_{mn}m)$ is fixed after the first optimization. Thus we can regard all these terms nodes $m$ in $Y$ as bias.

Moreover, what is more important is that,  for $m\in Y_+$, the pre-optimization will obtain its maximal value, and for $m\in Y_-$, this will get its minimal value. And for $n\in X_+$ or $n=z$, it will receive $Appx(W_{mn}m)$'s maximal values as bias either $m\in Y_+$ (then $W_{mn}>0$) or $m\in Y_-$ (then $W_{mn}<0$). Similar for $n\in X_-$. The exception is $n$ not in $Z$. But if $n\notin Z$, then any $m<n$ cannot in $Y$ because $Y\subseteq Z$.   Hence, if $n$ is not in $Z$, then $n$ receives no bias from $Y$.

Therefore, after the pre-optimization, for $z$ and every node $n\in X_+$, they receive nodes $m$ in $Y$ as bias, and this part of bias reaches its maximal. For every node $n\in X_-$,  they receive nodes $m$ in $Y$ as bias, and this part of bias reaches its minimal.

Let $I$ be the set of input nodes of $Y$, and $J$ be the set of rest input nodes.
After optimization of $Y$, the input values for $I$ are fixed, and therefore values of all nodes in $Y$.
Consider the DNN $D'$ with input nodes $J$, where all the 
edges from $m \in Y$ to $n$ are replaced by a bias for $n$ (we sum all these bias for a neuron $n$).
We denote $\val'$ values of nodes in this DNN $D'$. All the ReLU nodes of $D'$ are in $X$, by definition of $X$.
Thus $\MILP_X$ will be fully accurate in $D'$. Let $\beta'$ denote its upper bound. By the discussion earlier, we know that $\beta'=\max(z)$.

Similarly, we can prove $\alpha=\min(z)$. This concludes the proof of Theorem \ref{th2}

	\begin{proof}
	By contradiction, assume that there exist nodes $m, n^+ \in Y_+$ and  $n^
	- \in Y_-$, such that there are two non-zero weight paths $\rho_1$ from $m$ to $n^+$ and $\rho_2$ from $n$ to $n^-$ Assume first that the weights of $\rho_1,\rho_2$ have the same sign, 
	in which case it gives us a path $\rho_2 \rho_3$  from $m$ to $z$ of opposite sign than $\rho_1$. Applying Lemma \ref{lemma10}, we have prefixes $\rho'_1,\rho'_2$ of $\rho_1,\rho_2$ ending in the same state $m'$, and with weights of opposite sign.
	So we are left with path from $m'$ to $Y_-,Y_+$ of opposite sign. 
	We then apply Lemma \ref{lemma11}, inductively then applying Lemma \ref{lemma10} on the suffices etc. We obtain a contradiction to the existence of $m$ with $m<n^+$ and $m<n^-$. \qed
	\end{proof}
